---
title: "Estimation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 8,
  fig.height = 8,
  message = FALSE,
  warning = FALSE
)
library(ggplot2); library(dplyr); library(tidyr)
data.table::setDTthreads(1L)
options(dplyr.summarise.inform = FALSE, scipen = 999, digits = 5)
theme_set(theme_bw(base_size = 12))
```

> **Notation.** For symbol definitions, see the [notation vignette](notation.html).

## Overview

This vignette shows how to estimate DID, TD, and NTD using `childpen`. All estimators are computed jointly by `multiple_treatment_group_analysis()`. The function returns one row per treatment group $\times$ event time $\times$ estimand $\times$ method combination.

## Simulate data and run estimation

```{r draw_data}
library(childpen)

set.seed(42)
data <- simulate_data(n_individuals = 2000, treatment_groups = 26:32)
head(data)
```

```{r estimation}
res = multiple_treatment_group_analysis(data = data,
                                  treatment_groups = 26:28,
                                  periods_post = 3,
                                  periods_pre = NULL,
                                  max_age = 40,
                                  min_age = 20,
                                  pre = 1,
                                  verbose = FALSE
                                  )
```

## Simulation truth

The simulation provides potential outcomes so we can benchmark every estimand against its true value.

```{r truth}
PO_tidy <- data |>
  mutate(event_time = age - D) |>
  rename(d = D) |>
  group_by(female, d, event_time) |>
  summarize(APO_obs = mean(Y), APO = mean(Y_inf)) |>
  mutate(ATE = APO_obs - APO,
         theta = ATE / APO) |>
  filter(event_time %in% 0:3, d %in% 26:28)
```

## DID

DID estimates gender-specific effects using closest-not-yet-treated control groups. For treatment group $d$ and target age $a$, the control group is $d^\prime = a + 1$. The DID counterfactual is

$$\delta_{\mathrm{APO}}(g,d,d^\prime,a)=\mathbb{E}[Y_{d-1}\mid G=g, D=d]+\mathbb{E}[Y_a-Y_{d-1}\mid G=g, D=d^\prime]$$

and the DID ATE and normalised effect are

$$\delta_{\mathrm{ATE}}(g,d,d^\prime,a)=\mathbb{E}[Y_a\mid G=g, D=d]-\delta_{\mathrm{APO}}(g,d,d^\prime,a), \qquad \delta_{\theta}=\delta_{\mathrm{ATE}}/\delta_{\mathrm{APO}}.$$

```{r did_truth}
obs_tidy = data |>
  mutate(event_time = age - D) |>
  rename(d = D) |>
  group_by(female, d, event_time) |>
  summarize(est = mean(Y)) |>
  mutate(estimand = "APO", method = "Observed") |>
  filter(event_time %in% -3:3, d %in% 26:28)

PO_long = PO_tidy |>
  select(-APO_obs) |>
  pivot_longer(
    !c(female, d, event_time), names_to = "estimand", values_to = "est"
  ) |>
  mutate(method = "Truth")

plot_data = res |>
  filter(method %in% c("DID_Female", "DID_Male")) |>
  mutate(female = ifelse(method == "DID_Female", 1, 0)) |>
  select(d, event_time, estimand, female, est, ci_l, ci_h) |>
  mutate(method = "DID") |>
  dplyr::bind_rows(obs_tidy, PO_long)
```

### Females

```{r did_female}
plot_data |>
  filter(female == 1) |>
  ggplot(aes(x = event_time, y = est, ymin = ci_l, ymax = ci_h, color = method, fill = method)) +
  geom_ribbon(color = NA, alpha = .2) +
  geom_point() + geom_line() +
  facet_grid(cols = vars(d), rows = vars(estimand), scales = "free") +
  labs(x = "Event Time", y = "Estimate +/- 95% CI", color = "Method", fill = "Method") +
  theme(legend.position = "bottom")
```

### Males

```{r did_male}
plot_data |>
  filter(female == 0) |>
  ggplot(aes(x = event_time, y = est, ymin = ci_l, ymax = ci_h, color = method, fill = method)) +
  geom_ribbon(color = NA, alpha = .2) +
  geom_point() + geom_line() +
  facet_grid(cols = vars(d), rows = vars(estimand), scales = "free") +
  labs(x = "Event Time", y = "Estimate +/- 95% CI", color = "Method", fill = "Method") +
  theme(legend.position = "bottom")
```

## TD

TD estimates the gender gap in treatment effects in levels:

$$\mathrm{TD}(d, d^\prime, a) = \delta_{\mathrm{ATE}}(f, d, d^\prime, a) - \delta_{\mathrm{ATE}}(m, d, d^\prime, a).$$

```{r td_truth}
truth_td <- PO_tidy |>
  select(female, d, event_time, ATE) |>
  pivot_wider(names_from = female, values_from = ATE, names_prefix = "ATE_f") |>
  rename(ATE_female = ATE_f1, ATE_male = ATE_f0) |>
  mutate(est = ATE_female - ATE_male,
         method = "TD", estimand = "ATE",
         ci_l = NA_real_, ci_h = NA_real_,
         truth = TRUE)
```

```{r td_plot}
plot_td <- res |>
  filter(method == "TD") |>
  mutate(truth = FALSE) |>
  select(d, event_time, estimand, method, est, ci_l, ci_h, truth) |>
  bind_rows(truth_td |> select(d, event_time, estimand, method, est, ci_l, ci_h, truth)) |>
  mutate(label = ifelse(truth, "Truth", "TD"))

plot_td |>
  ggplot(aes(x = event_time, y = est, ymin = ci_l, ymax = ci_h,
             color = label, fill = label)) +
  geom_ribbon(color = NA, alpha = 0.2) +
  geom_point() + geom_line() +
  facet_grid(cols = vars(d), rows = vars(estimand), scales = "free") +
  labs(x = "Event Time", y = "Estimate +/- 95% CI",
       color = "Method", fill = "Method",
       title = "TD: ATE(female) - ATE(male)") +
  theme(legend.position = "bottom")
```

## NTD

NTD produces two estimands: `NTD_Conv` (conventional) and `NTD_New`.

**NTD_Conv** is the gap in normalised effects — the conventional child penalty:

$$\mathrm{NTD\_Conv}(d, d^\prime, a) = \delta_{\theta}(f, d, d^\prime, a) - \delta_{\theta}(m, d, d^\prime, a).$$

**NTD_New** measures the effect of parenthood on the gender earnings ratio $\rho$:

$$\mathrm{NTD\_New}(d, d^\prime, a) = \frac{\mathbb{E}[Y_a \mid f, D=d]}{\mathbb{E}[Y_a \mid m, D=d]} - \frac{\delta_{\mathrm{APO}}(f, d, d^\prime, a)}{\delta_{\mathrm{APO}}(m, d, d^\prime, a)} = \Delta\rho.$$

```{r ntd_truth}
truth_ntd_conv <- PO_tidy |>
  select(female, d, event_time, theta) |>
  pivot_wider(names_from = female, values_from = theta, names_prefix = "th_f") |>
  rename(th_female = th_f1, th_male = th_f0) |>
  mutate(est = th_female - th_male,
         method = "NTD_Conv", estimand = "theta",
         ci_l = NA_real_, ci_h = NA_real_,
         truth = TRUE)

truth_ntd_new <- PO_tidy |>
  select(female, d, event_time, APO_obs, APO) |>
  pivot_wider(names_from = female,
              values_from = c(APO_obs, APO),
              names_sep = "_f") |>
  rename(Y_female = APO_obs_f1, Y_male = APO_obs_f0,
         APO_female = APO_f1, APO_male = APO_f0) |>
  mutate(est = (Y_female / Y_male) - (APO_female / APO_male),
         method = "NTD_New", estimand = "Delta_rho",
         ci_l = NA_real_, ci_h = NA_real_,
         truth = TRUE)

truth_all <- bind_rows(truth_ntd_conv, truth_ntd_new) |>
  select(d, event_time, estimand, method, est, ci_l, ci_h, truth)
```

### NTD_Conv

```{r ntd_conv_plot}
plot_ntd_conv <- res |>
  filter(method == "NTD_Conv") |>
  mutate(truth = FALSE) |>
  select(d, event_time, estimand, method, est, ci_l, ci_h, truth) |>
  bind_rows(truth_all |> filter(method == "NTD_Conv")) |>
  mutate(label = ifelse(truth, "Truth", "NTD_Conv"))

plot_ntd_conv |>
  ggplot(aes(x = event_time, y = est, ymin = ci_l, ymax = ci_h,
             color = label, fill = label)) +
  geom_ribbon(color = NA, alpha = 0.2) +
  geom_point() + geom_line() +
  facet_grid(cols = vars(d), rows = vars(estimand), scales = "free") +
  labs(x = "Event Time", y = "Estimate +/- 95% CI",
       color = "Method", fill = "Method",
       title = "NTD_Conv: theta(female) - theta(male)") +
  theme(legend.position = "bottom")
```

### NTD_New

```{r ntd_new_plot}
plot_ntd_new <- res |>
  filter(method == "NTD_New") |>
  mutate(truth = FALSE) |>
  select(d, event_time, estimand, method, est, ci_l, ci_h, truth) |>
  bind_rows(truth_all |> filter(method == "NTD_New")) |>
  mutate(label = ifelse(truth, "Truth", "NTD_New"))

plot_ntd_new |>
  ggplot(aes(x = event_time, y = est, ymin = ci_l, ymax = ci_h,
             color = label, fill = label)) +
  geom_ribbon(color = NA, alpha = 0.2) +
  geom_point() + geom_line() +
  facet_grid(cols = vars(d), rows = vars(estimand), scales = "free") +
  labs(x = "Event Time", y = "Estimate +/- 95% CI",
       color = "Method", fill = "Method",
       title = expression(paste("NTD_New: ", Delta, rho, " (effect on gender earnings ratio)"))) +
  theme(legend.position = "bottom")
```

## Validation tests

For a discussion on how to perform validation tests, and what type of tests are correct for this type of estimation, see the [validation tests vignette](validation_tests.html).
