---
title: "NTD-identification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NTD-identification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 8,
  fig.height = 8,
  message = FALSE,
  warning = FALSE
)
library(ggplot2); library(dplyr); library(tidyr)
options(dplyr.summarise.inform = FALSE, scipen = 999, digits = 5)
theme_set(theme_bw(base_size = 12))
```

## Overview

The aim of this vignette is to create intuition on the normalized triple differences (NTD) identification framework discussed in the paper. NTD is the underlying identification framework for normalized event studies, a common empirical strategy in the child penalty literature. 

**This vignette demonstrates:**

1. The identification assumption in NTD
2. The bias that emerges from normalizing

We'll use a simple data generating process (DGP) to visualize everything. You are encouraged to play with the code to see how the bias changes under different parameters. 

---

## The DGP

We create a DGP with:

- Two genders: women and men (with men earning more)
- Two treatment cohorts: early (D=25) and late (D=30)  
- Linear earnings growth, with late cohorts have steeper slopes
- Earnings penalties from childbirth

The key feature: when taking ratios within gender, the difference between men and women is netted out

```{r dgp, class.source = "fold-show"}
# Parameters
baseline    <- 500
ratio_bias  <- 3    # late vs early slope multiplier
ratio_apos  <- 3    # men vs women slope multiplier
ages        <- 20:30
d_early     <- 25
d_late      <- 30

# Counterfactual earning slopes
s_f_25 <- baseline
s_f_30 <- baseline * ratio_bias
s_m_25 <- baseline * ratio_apos
s_m_30 <- baseline * ratio_apos * ratio_bias

# Treatment effects (ATT)
att_f  <- -6000
att_m  <- -2000

# Build each group
make_group <- function(sex, D, slope, att_level) {
  tibble(
    age    = ages,
    female = sex,
    D      = D,
    y_0    = slope * age,           # counterfactual (no treatment)
    y_1    = y_0 + ifelse(age >= D, att_level, 0),
    y      = y_1                     # observed
  )
}

data <- bind_rows(
  make_group(1, d_early, s_f_25, att_f),
  make_group(1, d_late,  s_f_30, att_f),
  make_group(0, d_early, s_m_25, att_m),
  make_group(0, d_late,  s_m_30, att_m)
)
```

## Verifying the NTD Assumption Holds

Recall, that NTD assumea that normalized violations of parallel trends are equal across genders.

Before visualizing the bias, let's verify that this DGP actually satisfies this assumption.

### Parallel trends violation in levels

First, let's calculate the parallel trend violation in levels - the difference in counterfactual trends between early and late treated within gender:

```{r verify_pt_levels}
# Calculate trend from pre-treatment (age D-1) to each subsequent age
trends <- function(g, d, a) {
  pre <- data |> filter(D == d, female == g, age == d-1) |> pull(y_0)
  post <- data |> filter(D == d, female == g, age == a) |> pull(y_0)
  post - pre
}

# Difference in trends between early and late cohorts
diff_trends <- function(g, a) trends(g, 25, a) - trends(g, 30, a)

expand_grid(
  female = c(0, 1),
  age    = ages
) |>
  mutate(
    diff_trend = purrr::map2_dbl(female, age, diff_trends),
    gender = factor(if_else(female == 1, "Women", "Men"), levels = c("Women", "Men"))
  ) |> 
  ggplot(aes(x = age, y = diff_trend, color = gender)) + 
  geom_line(linewidth = 1.1) + 
  labs(
    title = "Parallel Trends Violation in Levels by Gender",
    x = "Age", 
    y = "Difference in counterfactual trends", 
    color = "Gender"
  ) +
  theme(legend.position = "bottom")
```

The parallel trends violations in levels are **not equal** across gender. Men's violations are much larger because they have higher  earnings. This violates both difference-in-differences (DID), whichassumes no violation, and triple differences (TD), which assumes equal violations in levels across genders.

### Normalized parallel trends violation

Now let's normalize by dividing by the average potential earnings (APO) in the absence of treatment:

```{r verify_normalized_pt}
# Get counterfactual earnings for early cohort (APO)
get_apo <- function(g, a) {
  data |>
    dplyr::filter(D == 25, female == g, age == a) |>
    dplyr::pull(y_0)
}

expand_grid(
  female = c(0, 1),
  age    = ages
) |>
  mutate(
    diff_trend = purrr::map2_dbl(female, age, diff_trends),
    y_0 = purrr::map2_dbl(female, age, get_apo),
    norm_diff_trend = diff_trend / y_0,
    gender = factor(if_else(female == 1, "Women", "Men"), levels = c("Women", "Men"))
  ) |> 
  ggplot(aes(x = age, y = norm_diff_trend, color = gender)) + 
  geom_line(linewidth = 1.1) + 
  labs(
    title = "Normalized Parallel Trends Violation",
    x = "Age", 
    y = "Normalized difference in counterfactual trends", 
    color = "Gender"
  ) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme(legend.position = "bottom")
```

**Taekaway:** The normalized parallel trends violations are identical for men and women at every age. This is exactly what the NTD  identification assumption requires.

**Why does this work?** By construction, late cohort slopes are a constant multiple of early slopes (within gender), and men's slopes are a constant multiple of women's slopes (within cohort). These constant multiples for men cancel out when we take ratios, ensuring the normalized PT violations are equal across gender.

--

## The Setup - Counterfactuals and Observed Data

Let's start by seeing what we observe vs. the true counterfactuals:

```{r fig1_setup, fig.height=4}
plot_df <- bind_rows(
  # Early cohort: observed and counterfactual
  data %>%
    filter(D == d_early) %>%
    pivot_longer(cols = c(y_0, y), names_to = "series", values_to = "value"),
  # Late cohort: counterfactual only (not yet treated at these ages for early cohort)
  data %>%
    filter(D == d_late) %>%
    transmute(age, female, D, series = "y_0", value = y_0)
) %>%
  mutate(
    gender = factor(if_else(female == 1, "Women", "Men"), levels = c("Women", "Men")),
    cohort = factor(if_else(D == d_early, "Early (D=25)", "Late (D=30)"),
                    levels = c("Early (D=25)", "Late (D=30)")),
    series_label = case_when(
      cohort == "Early (D=25)" & series == "y"   ~ "Early — Observed",
      cohort == "Early (D=25)" & series == "y_0" ~ "Early — Counterfactual",
      cohort == "Late (D=30)"  & series == "y_0" ~ "Late — Counterfactual"
    ),
    linetype = if_else(series_label == "Early — Observed", "solid", "dashed"),
    color = case_when(
      series_label == "Early — Observed"       ~ "Obs.",
      series_label == "Early — Counterfactual" ~ "CF Early",
      TRUE                                     ~ "CF Late"
    )
  )

ggplot(plot_df, aes(x = age, y = value, group = series_label)) +
  geom_line(aes(linetype = linetype, color = color), linewidth = 1.1) +
  facet_wrap(~ gender, nrow = 1) +
  scale_linetype_identity() +
  scale_color_manual(
    values = c("Obs." = "#1f77b4", "CF Early" = "#9ecae1", "CF Late" = "#fdae6b")
  ) +
  labs(
    title = "Figure 1: Observed vs. Counterfactual Earnings",
    x = "Age", y = "Earnings", color = NULL
  ) +
  theme(legend.position = "bottom")
```

- The solid blue line shows what we observe for the early cohort after treatment at age 25
- Light blue dashed: what would have happened without treatment (early cohort)
- Orange dashed: counterfactual for late treatment group
- The gap between solid and light blue dashed = average treatment effect (ATE)

**The problem:** We don't observe the light blue line. We need to impute it somehow.

## The DID Approach - Using Late Cohort as Control

DID uses the trend of the late cohort to impute the counterfactual for the early cohort:

```{r fig2_did}
# Calculate DID-imputed counterfactual
did_cf <- bind_rows(lapply(c(1, 0), function(g) {
  # Early cohort's pre-treatment level
  early_pre <- data %>%
    filter(female == g, D == d_early, age == d_early - 1) %>%
    pull(y)
  
  # Late cohort trend
  late_cf <- data %>% filter(female == g, D == d_late) %>% select(age, y_0)
  late_pre <- late_cf %>% filter(age == d_early - 1) %>% pull(y_0)
  
  # DID imputation: shift late trend to match early pre-treatment level
  tibble(
    age    = late_cf$age,
    female = g,
    D      = d_early,
    series = "y_cf_did",
    value  = early_pre + (late_cf$y_0 - late_pre)
  )
})) %>%
  mutate(
    gender = factor(if_else(female == 1, "Women", "Men"), levels = c("Women", "Men")),
    cohort = factor("Early (D=25)", levels = c("Early (D=25)", "Late (D=30)")),
    series_label = "Early — DID-imputed CF",
    linetype = "dotdash",
    color = "CF DID"
  )

plot_df_plus <- bind_rows(plot_df, did_cf)

ggplot(plot_df_plus, aes(x = age, y = value, group = series_label)) +
  geom_line(aes(linetype = linetype, color = color), linewidth = 1.1) +
  facet_wrap(~ gender, nrow = 1) +
  scale_linetype_identity() +
  scale_color_manual(
    values = c(
      "Obs." = "#1f77b4", "CF Early" = "#9ecae1", 
      "CF Late" = "#fdae6b", "CF DID" = "#2ca02c"
    )
  ) +
  labs(
    title = "Figure 2: DID-Imputed Counterfactual",
    subtitle = "Green line = what DID thinks would have happened without treatment",
    x = "Age", y = "Earnings", color = NULL
  ) +
  theme(legend.position = "bottom")
```

The green line (DID imputed counterfactual) doesn't match the light blue line (true counterfactual). This difference is bias due to parallel trend violations. It happens because the late cohort has a steeper slope for counterfactual earnings.

## Decomposing DID Estimates into Truth and Bias

Let's break down what DID gives us. First, in the above figure, we have:

- DID APO (green line)
- Truth APO (light blue line)
- DID ATE (difference between green and light blue lines)
- Truth ATE (difference between blue and light blue lines)
- PT Bias (difference between green and light blue lines)

For each age, we can decompose:

- DID APO = True APO + PT Bias
- DID ATE = True ATE - PT Bias (DID ATE is difference between orange and green lines, so PT bias comes here in minus)

Below, I calculate the APO and ATE under DID. And then show the decomposition of these value, by age and gender, into truth and bias. 

```{r fig3_decomposition}
# Calculate true APO and ATE
early_true_age <- data %>%
  filter(D == d_early) %>%
  group_by(female, age) %>%
  summarise(APO_true = y_0, APO_obs = y, .groups = "drop") |> 
  mutate(ATE_true = APO_obs - APO_true)

# Calculate DID APO
early_did_age <- did_cf %>%
  group_by(female, age) %>%
  summarise(APO_did = value, .groups = "drop")

# Combine and calculate bias
summary_age <- early_true_age %>%
  left_join(early_did_age, by = c("female", "age")) %>%
  mutate(
    PT_bias = APO_did - APO_true,  # parallel trends bias
    ATE_did = APO_obs - APO_did,  # DID ATE = truth + bias
    gender  = factor(if_else(female == 1, "Women", "Men"), levels = c("Women", "Men"))
  )

# Prepare stacked bar chart data
apo_stack_age <- summary_age %>%
  transmute(
    gender, age = factor(age),
    measure = factor("APO", levels = c("APO", "ATE")),
    total = APO_did, comp_true = APO_true, comp_bias = PT_bias
  ) %>%
  pivot_longer(c(comp_true, comp_bias), names_to = "component", values_to = "value") %>%
  mutate(component = if_else(component == "comp_true", "Truth", "PT bias"))

ate_stack_age <- summary_age %>%
  transmute(
    gender, age = factor(age),
    measure = factor("ATE", levels = c("APO", "ATE")),
    total = ATE_did, comp_true = ATE_true, comp_bias = -PT_bias
  ) %>%
  pivot_longer(c(comp_true, comp_bias), names_to = "component", values_to = "value") %>%
  mutate(component = if_else(component == "comp_true", "Truth", "PT bias"))

bar_age_df <- bind_rows(apo_stack_age, ate_stack_age)

ggplot(bar_age_df, aes(x = age, y = value, fill = component)) +
  geom_col(width = 0.7) +
  facet_grid(rows = vars(measure), cols = vars(gender), scales = "free_y") +
  scale_fill_manual(
    values = c("Truth" = "#1f77b4", "PT bias" = "#fdae6b"),
    breaks = c("Truth", "PT bias")
  ) +
  labs(
    title = "Decomposing DID into effect and bias",
    subtitle = "Top row: DID APO | Bottom row: DID ATE",
    x = "Age", y = NULL, fill = NULL
  ) +
  theme(legend.position = "bottom")
```

This is not suprising: when counterfactual trends are not equal, DID does not identify the APO, and hence doesnt identify the ATE. 

## The Bias from Normalizing

Now for the key question: **What happens when we normalize?**

In NTD, we compute: $$\delta_{\theta} = \frac{\text{DID ATE}}{\text{DID APO}}$$

This is meant to estimate the true ratio: $$\theta = \frac{\text{True ATE}}{\text{True APO}}$$

### Step 1: The PT Bias Component

Note that  $$\delta_{\theta} = \frac{\text{DID ATE}}{\text{DID APO}}=\frac{\text{TRUE ATE} - \text{PT Bias}}{\text{DID APO}}=\frac{\text{TRUE ATE}}{\text{DID APO}}-\frac{\text{PT Bias}}{\text{DID APO}}$$

We will now analyze $\frac{\text{TRUE ATE}}{\text{DID APO}}+\frac{\text{PT Bias}}{\text{DID APO}}$. Since this is a sum, will go one by one. Start with $-\frac{\text{PT Bias}}{\text{DID APO}}$.

```{r fig4_pt_ratio}
pt_ratio_df <- summary_age %>%
  transmute(
    gender, age = factor(age),
    measure = factor("Ratios", levels = c("APO","ATE","Ratios")),
    component = "PT / DID APO",
    value = -PT_bias / APO_did
  )

bar_age_three <- bind_rows(
  apo_stack_age %>% mutate(measure = factor("APO", levels = c("APO","ATE","Ratios"))),
  ate_stack_age %>% mutate(measure = factor("ATE", levels = c("APO","ATE","Ratios"))),
  pt_ratio_df
)

ggplot(bar_age_three, aes(x = age, y = value, fill = component)) +
  geom_col(width = 0.7) +
  facet_grid(rows = vars(measure), cols = vars(gender), scales = "free_y") +
  scale_fill_manual(
    values = c(
      "Truth" = "#1f77b4", "PT bias" = "#fdae6b", "PT / DID APO" = "#99d8c9"
    )
  ) +
  labs(
    title = "Figure 4: Normalizing the PT Bias",
    subtitle = "Bottom row shows PT bias / DID APO",
    x = "Age", y = NULL, fill = NULL
  ) +
  theme(legend.position = "bottom")
```

**Takeaway:** The bottom row (PT / DID APO) is **identical for men and women** at every age. Hence, under NTD, differencing across gender cancels out this bias parameter.

We now continue with the second component

### Step 2: The Effect Component

Now let's add the second part of the sum, $\frac{\text{ATE}}{\text{DID APO}}$.

```{r fig5_full_ratio}
ratios_df <- summary_age %>%
  transmute(
    gender, age = factor(age),
    measure = factor("Ratios", levels = c("APO","ATE","Ratios")),
    `PT ratio`  = -PT_bias / APO_did,
    `ATE ratio` = ATE_true / APO_did
  ) %>%
  pivot_longer(
    cols = c(`PT ratio`, `ATE ratio`), 
    names_to = "component", 
    values_to = "value"
  )

bar_age_three2 <- bind_rows(
  apo_stack_age %>% mutate(measure = factor("APO", levels = c("APO","ATE","Ratios"))),
  ate_stack_age %>% mutate(measure = factor("ATE", levels = c("APO","ATE","Ratios"))),
  ratios_df
)

ggplot(bar_age_three2, aes(x = age, y = value, fill = component)) +
  geom_col(width = 0.7) +
  facet_grid(rows = vars(measure), cols = vars(gender), scales = "free_y") +
  scale_fill_manual(
    values = c(
      "Truth" = "#1f77b4", "PT bias" = "#fdae6b",
      "PT ratio" = "#99d8c9", "ATE ratio" = "#2ca25f"
    )
  ) +
  labs(
    title = "Figure 5: Complete Decomposition of the Normalized Ratio",
    subtitle = "Bottom row: Light teal (drops out when differencing) + Dark green (what remains)",
    x = "Age", y = NULL, fill = NULL
  ) +
  theme(legend.position = "bottom")
```

The bottom row shows:

- Light teal (PT / DID APO): Same across gender $\rightarrow$ cancels when we difference between gender
- Dark green (ATE / DID APO): Different across gender 

Now we need to examine, if we difference (ATE / DID APO) between gender, does this capture the gender gap in normalized effects?

## The Multiplicative Bias

Here's where it gets interesting. The (ATE / DID APO) (dark green in above figure) can be rewritten to reveal a multiplicative bias:

$$\frac{\text{True ATE}}{\text{DID APO}} = \frac{\text{True APO} \cdot \theta}{\text{DID APO}} = \theta \cdot \frac{\text{True APO}}{\text{DID APO}}$$

where, recall, $\theta = \text{True ATE} / \text{True APO}$ is the causal estimand we are interested in. Or, at least, the gender gap in $\theta$.

Adding and subtracting $\theta$, and using the definition of DID APO = True APO + PT Bias, we can re-write the above as
\[
\begin{aligned}
\frac{\text{True ATE}}{\text{DID APO}} &= \theta \frac{\text{True APO}}{\text{DID APO}} \\
&= \theta - \theta + \theta \frac{\text{True APO}}{\text{DID APO}} \\
&= \theta + \theta \left(\frac{\text{True APO}}{\text{DID APO}} - 1\right) \\
&= \theta + \theta \left(\frac{\text{True APO - True APO - PT Bias}}{\text{DID APO}}\right) \\
&= \theta - \theta \left(\frac{\text{PT Bias}}{\text{DID APO}}\right)
\end{aligned}
\]

The second term is **multiplicative bias** - the true effect gets scaled by the ratio of PT bias to DID APO. As we discussed above, this quantity $\frac{\text{PT bias}}{\text{DID APO}}$ is constant between genders. So both genders treatment effects are scaled by the same factor. 

```{r fig6_multiplicative}

## sanity check
summary_age %>%
  transmute(
    gender, age = factor(age),
    measure = factor("Ratios", levels = c("APO","ATE","Ratios")),
    `ATE ratio` = ATE_true / APO_did,
    `ATE ratio check` = (ATE_true / APO_true) + (ATE_true / APO_true) * (PT_bias / APO_did)
  )

theta_df <- summary_age %>%
  transmute(
    gender, age = factor(age),
    APO_true, ATE_true, APO_did, PT_bias,
    theta_true = ATE_true / APO_true,
    bias_A = -PT_bias / APO_did,
    bias_B = - theta_true * (PT_bias / APO_did)
  )

theta_stack <- bind_rows(
  theta_df %>% transmute(
    gender, age, measure = factor("Ratios", levels = c("APO","ATE","Ratios")),
    component = "Truth", value = theta_true
  ),
  theta_df %>% transmute(
    gender, age, measure = factor("Ratios", levels = c("APO","ATE","Ratios")),
    component = "Bias A", value = bias_A
  ),
  theta_df %>% transmute(
    gender, age, measure = factor("Ratios", levels = c("APO","ATE","Ratios")),
    component = "Bias B", value = bias_B
  )
)

bar_age_three3 <- bind_rows(
  apo_stack_age %>% mutate(measure = factor("APO", levels = c("APO","ATE","Ratios"))),
  ate_stack_age %>% mutate(measure = factor("ATE", levels = c("APO","ATE","Ratios"))),
  theta_stack
)

ggplot(bar_age_three3, aes(x = age, y = value, fill = component)) +
  geom_col(width = 0.7) +
  facet_grid(rows = vars(measure), cols = vars(gender), scales = "free_y") +
  scale_fill_manual(
    values = c(
      "Truth" = "#1f77b4", "PT bias" = "#fdae6b",
      "Bias A" = "#99d8c9", "Bias B" = "#2ca25f"
    ),
    breaks = c("Truth", "PT bias", "Bias A", "Bias B"),
    labels = c("Truth", "PT bias", "PT / DID APO", "Truth × Bias")
  ) +
  labs(
    title = "Figure 6: The Multiplicative Bias Structure",
    subtitle = "Bottom row: Blue (true θ) + Teal (drops out) + Green (multiplicative bias)",
    x = "Age", y = NULL, fill = NULL
  ) +
  theme(legend.position = "bottom")
```

**Bottom row interpretation:**

- **Blue (Truth)**: The true ratio $\theta$ we want to estimate
- **Teal (Bias A)**: PT / DID APO - drops out when differencing across gender
- **Green (Bias B)**: $\theta \times \frac{\text{PT bias}}{\text{DID APO}}$ - is not equal between gender, since $\theta$ is allowed to be different between gender $\rightarrow$ hence bias remains when differencing

Note that the size of bias (in Bias B in the above plot) depends on relation of PT bias to APO (where this relation is constant by gender under NTD). 
