---
title: "NTD-identification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NTD-identification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 10,
  fig.height = 5.5,
  dpi = 120,
  message = FALSE,
  warning = FALSE
)
library(ggplot2); library(dplyr); library(tidyr)
options(dplyr.summarise.inform = FALSE, scipen = 999, digits = 5)
theme_set(theme_bw(base_size = 12))
```

## Overview

Normalized event studies are widely used in labor economics to compare treatment effects across groups with different baseline levels (e.g., comparing earnings penalties from childbirth for men and women). The standard approach divides effects by earnings levels to make them comparable.

**This vignette demonstrates:**

1. What identification assumption makes normalized comparisons valid
2. The bias structure that emerges from normalizing
3. How this bias behaves when comparing across groups

We'll use a simple data generating process (DGP) to visualize exactly where bias comes from and why it matters.

## Setup: A DGP where normalized parallel trends holds

We create a world with:

- Two genders: women and men (with men earning more)
- Two treatment cohorts: early (D=25) and late (D=30)  
- Linear earnings growth (late cohorts have steeper slopes)
- Treatment effects (earnings penalties from childbirth)

The key design feature: **ratios are preserved** such that normalized parallel trends violations are equal across gender.

```{r dgp, class.source = "fold-show"}
# Parameters
baseline    <- 500
ratio_bias  <- 3    # late vs early slope multiplier
ratio_apos  <- 3    # men vs women slope multiplier
ages        <- 20:30
d_early     <- 25
d_late      <- 30

# Counterfactual earning slopes
s_f_25 <- baseline
s_f_30 <- baseline * ratio_bias
s_m_25 <- baseline * ratio_apos
s_m_30 <- baseline * ratio_apos * ratio_bias

# Treatment effects (ATT)
att_f  <- -6000
att_m  <- -2000

# Build each group
make_group <- function(sex, D, slope, att_level) {
  tibble(
    age    = ages,
    female = sex,
    D      = D,
    y_0    = slope * age,           # counterfactual (no treatment)
    y_1    = y_0 + ifelse(age >= D, att_level, 0),
    y      = y_1                     # observed
  )
}

data <- bind_rows(
  make_group(1, d_early, s_f_25, att_f),
  make_group(1, d_late,  s_f_30, att_f),
  make_group(0, d_early, s_m_25, att_m),
  make_group(0, d_late,  s_m_30, att_m)
)
```

## Verifying the NTD Assumption Holds

Before visualizing the bias, let's verify that this DGP actually satisfies the normalized parallel trends assumption.

### Parallel trends violation in levels

First, let's calculate the parallel trend violation in levels - the difference in counterfactual trends between early and late treated within gender:

```{r verify_pt_levels, fig.height=4}
# Calculate trend from pre-treatment (age D-1) to each subsequent age
trends <- function(g, d, a) {
  pre <- data |> filter(D == d, female == g, age == d-1) |> pull(y_0)
  post <- data |> filter(D == d, female == g, age == a) |> pull(y_0)
  post - pre
}

# Difference in trends between early and late cohorts
diff_trends <- function(g, a) trends(g, 25, a) - trends(g, 30, a)

expand_grid(
  female = c(0, 1),
  age    = ages
) |>
  mutate(
    diff_trend = purrr::map2_dbl(female, age, diff_trends),
    gender = factor(if_else(female == 1, "Women", "Men"), levels = c("Women", "Men"))
  ) |> 
  ggplot(aes(x = age, y = diff_trend, color = gender)) + 
  geom_line(linewidth = 1.1) + 
  labs(
    title = "Parallel Trends Violation in Levels",
    subtitle = "Different across gender → standard parallel trends fails",
    x = "Age", 
    y = "Difference in counterfactual trends (dollars)", 
    color = "Gender"
  ) +
  theme(legend.position = "bottom")
```

**Key observation:** The parallel trends violations in levels are **not equal** across gender. Men's violations are much larger because they have higher baseline earnings. This violates standard parallel trends.

### Normalized parallel trends violation

Now let's normalize by dividing by the average potential earnings (APO) in the absence of treatment:

```{r verify_normalized_pt, fig.height=4}
# Get counterfactual earnings for early cohort (APO)
get_apo <- function(g, a) {
  data |>
    dplyr::filter(D == 25, female == g, age == a) |>
    dplyr::pull(y_0)
}

expand_grid(
  female = c(0, 1),
  age    = ages
) |>
  mutate(
    diff_trend = purrr::map2_dbl(female, age, diff_trends),
    y_0 = purrr::map2_dbl(female, age, get_apo),
    norm_diff_trend = diff_trend / y_0,
    gender = factor(if_else(female == 1, "Women", "Men"), levels = c("Women", "Men"))
  ) |> 
  ggplot(aes(x = age, y = norm_diff_trend, color = gender)) + 
  geom_line(linewidth = 1.1) + 
  labs(
    title = "Normalized Parallel Trends Violation",
    subtitle = "Identical across gender → NTD assumption holds!",
    x = "Age", 
    y = "Normalized difference (proportion of earnings)", 
    color = "Gender"
  ) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme(legend.position = "bottom")
```

**Critical result:** The normalized parallel trends violations are **identical** for men and women at every age. The lines overlap perfectly. This is exactly what the NTD (normalized triple differences) identification assumption requires.

**Why does this work?** By construction, late cohort slopes are a constant multiple of early slopes (within gender), and men's slopes are a constant multiple of women's slopes (within cohort). These constant multiples cancel out when we take ratios, ensuring the normalized PT violations are equal across gender.

## The Setup - Counterfactuals and Observed Data

Let's start by seeing what we observe vs. the true counterfactuals:

```{r fig1_setup, fig.height=4}
plot_df <- bind_rows(
  # Early cohort: observed and counterfactual
  data %>%
    filter(D == d_early) %>%
    pivot_longer(cols = c(y_0, y), names_to = "series", values_to = "value"),
  # Late cohort: counterfactual only (not yet treated at these ages for early cohort)
  data %>%
    filter(D == d_late) %>%
    transmute(age, female, D, series = "y_0", value = y_0)
) %>%
  mutate(
    gender = factor(if_else(female == 1, "Women", "Men"), levels = c("Women", "Men")),
    cohort = factor(if_else(D == d_early, "Early (D=25)", "Late (D=30)"),
                    levels = c("Early (D=25)", "Late (D=30)")),
    series_label = case_when(
      cohort == "Early (D=25)" & series == "y"   ~ "Early — Observed",
      cohort == "Early (D=25)" & series == "y_0" ~ "Early — Counterfactual",
      cohort == "Late (D=30)"  & series == "y_0" ~ "Late — Counterfactual"
    ),
    linetype = if_else(series_label == "Early — Observed", "solid", "dashed"),
    color = case_when(
      series_label == "Early — Observed"       ~ "Obs.",
      series_label == "Early — Counterfactual" ~ "CF Early",
      TRUE                                     ~ "CF Late"
    )
  )

ggplot(plot_df, aes(x = age, y = value, group = series_label)) +
  geom_line(aes(linetype = linetype, color = color), linewidth = 1.1) +
  facet_wrap(~ gender, nrow = 1) +
  scale_linetype_identity() +
  scale_color_manual(
    values = c("Obs." = "#1f77b4", "CF Early" = "#9ecae1", "CF Late" = "#fdae6b")
  ) +
  labs(
    title = "Figure 1: Observed vs. Counterfactual Earnings",
    x = "Age", y = "Earnings", color = NULL
  ) +
  theme(legend.position = "bottom")
```

**Key observations:**

- The solid blue line shows what we observe for the early cohort after treatment at age 25
- Light blue dashed: what would have happened without treatment (early cohort)
- Orange dashed: what would have happened for the late cohort (different slope!)
- The gap between solid and light blue dashed = true treatment effect

**The problem:** We don't observe the light blue line. We need to impute it somehow.

## The DID Approach - Using Late Cohort as Control

Standard difference-in-differences uses the late cohort to impute the counterfactual for the early cohort:

```{r fig2_did}
# Calculate DID-imputed counterfactual
did_cf <- bind_rows(lapply(c(1, 0), function(g) {
  # Early cohort's pre-treatment level
  early_pre <- data %>%
    filter(female == g, D == d_early, age == d_early - 1) %>%
    pull(y)
  
  # Late cohort trend
  late_cf <- data %>% filter(female == g, D == d_late) %>% select(age, y_0)
  late_pre <- late_cf %>% filter(age == d_early - 1) %>% pull(y_0)
  
  # DID imputation: shift late trend to match early pre-treatment level
  tibble(
    age    = late_cf$age,
    female = g,
    D      = d_early,
    series = "y_cf_did",
    value  = early_pre + (late_cf$y_0 - late_pre)
  )
})) %>%
  mutate(
    gender = factor(if_else(female == 1, "Women", "Men"), levels = c("Women", "Men")),
    cohort = factor("Early (D=25)", levels = c("Early (D=25)", "Late (D=30)")),
    series_label = "Early — DID-imputed CF",
    linetype = "dotdash",
    color = "CF DID"
  )

plot_df_plus <- bind_rows(plot_df, did_cf)

ggplot(plot_df_plus, aes(x = age, y = value, group = series_label)) +
  geom_line(aes(linetype = linetype, color = color), linewidth = 1.1) +
  facet_wrap(~ gender, nrow = 1) +
  scale_linetype_identity() +
  scale_color_manual(
    values = c(
      "Obs." = "#1f77b4", "CF Early" = "#9ecae1", 
      "CF Late" = "#fdae6b", "CF DID" = "#2ca02c"
    )
  ) +
  labs(
    title = "Figure 2: DID-Imputed Counterfactual",
    subtitle = "Green line = what DID thinks would have happened without treatment",
    x = "Age", y = "Earnings", color = NULL
  ) +
  theme(legend.position = "bottom")
```

**The key insight:** The green line (DID imputation) doesn't match the light blue line (truth). This difference is **parallel trends bias** - it happens because the late cohort has a steeper slope.

**Important property of this DGP:** When you divide the green line by the light blue dashed line, you get the **same ratio** for men and women at every age. This is an implication of thee normalized parallel trends assumption.

## Decomposing DID Estimates into Truth and Bias

Let's break down what DID gives us. For each age, we can decompose:

- **DID APO** (average potential outcome) = Truth - PT Bias  
- **DID ATE** (average treatment effect) = Truth + PT Bias

```{r fig3_decomposition, fig.height=5.5}
# Calculate true APO and ATE
early_true_age <- data %>%
  filter(D == d_early) %>%
  group_by(female, age) %>%
  summarise(APO_true = y_0, ATE_true = y - y_0, .groups = "drop")

# Calculate DID APO
early_did_age <- did_cf %>%
  group_by(female, age) %>%
  summarise(APO_did = value, .groups = "drop")

# Combine and calculate bias
summary_age <- early_true_age %>%
  left_join(early_did_age, by = c("female", "age")) %>%
  mutate(
    PT_bias = APO_true - APO_did,  # parallel trends bias
    ATE_did = ATE_true + PT_bias,  # DID ATE = truth + bias
    gender  = factor(if_else(female == 1, "Women", "Men"), levels = c("Women", "Men"))
  )

# Prepare stacked bar chart data
apo_stack_age <- summary_age %>%
  transmute(
    gender, age = factor(age),
    measure = factor("APO", levels = c("APO", "ATE")),
    total = APO_did, comp_true = APO_true, comp_bias = -PT_bias
  ) %>%
  pivot_longer(c(comp_true, comp_bias), names_to = "component", values_to = "value") %>%
  mutate(component = if_else(component == "comp_true", "Truth", "PT bias"))

ate_stack_age <- summary_age %>%
  transmute(
    gender, age = factor(age),
    measure = factor("ATE", levels = c("APO", "ATE")),
    total = ATE_did, comp_true = ATE_true, comp_bias = PT_bias
  ) %>%
  pivot_longer(c(comp_true, comp_bias), names_to = "component", values_to = "value") %>%
  mutate(component = if_else(component == "comp_true", "Truth", "PT bias"))

bar_age_df <- bind_rows(apo_stack_age, ate_stack_age)

ggplot(bar_age_df, aes(x = age, y = value, fill = component)) +
  geom_col(width = 0.7) +
  facet_grid(rows = vars(measure), cols = vars(gender), scales = "free_y") +
  scale_fill_manual(
    values = c("Truth" = "#1f77b4", "PT bias" = "#fdae6b"),
    breaks = c("Truth", "PT bias")
  ) +
  labs(
    title = "Decomposing DID",
    subtitle = "Top row: DID APO (DID imputed counterfactual earnings) | Bottom row: DID ATE (DID imputed treatment effect)",
    x = "Age", y = NULL, fill = NULL
  ) +
  theme(legend.position = "bottom")
```

**What this shows:**

- **APO row:** DID estimates of counterfactual earnings = True counterfactual (blue) minus PT bias (orange)
- **ATE row:** DID estimates of treatment effects = True effect (blue) plus PT bias (orange)

The bias has opposite signs because the DID counterfactual overshoots the truth (steeper slope).

## The Bias from Normalizing

Now for the key question: **What happens when we normalize?**

In normalized event studies, we compute: $$\delta_{\theta} = \frac{\text{DID ATE}}{\text{DID APO}}$$

This is meant to estimate the true ratio: $$\theta = \frac{\text{True ATE}}{\text{True APO}}$$

Let's see what bias emerges when we compute $\delta_{\theta}$.

### Step 1: The PT Bias Component

Note that  $$\delta_{\theta} = \frac{\text{DID ATE}}{\text{DID APO}}=\frac{\text{TRUE ATE} + \text{PT Bias}}{\text{DID APO}}=\frac{\text{TRUE ATE}}{\text{DID APO}}+\frac{\text{PT Bias}}{\text{DID APO}}$$

We will now analyze $\frac{\text{TRUE ATE}}{\text{DID APO}}+\frac{\text{PT Bias}}{\text{DID APO}}$. Since this is a sum, will go one by one. Start with $\frac{\text{PT Bias}}{\text{DID APO}}$.

```{r fig4_pt_ratio, fig.height=5.5}
pt_ratio_df <- summary_age %>%
  transmute(
    gender, age = factor(age),
    measure = factor("Ratios", levels = c("APO","ATE","Ratios")),
    component = "PT / DID APO",
    value = PT_bias / pmax(APO_true - PT_bias, .Machine$double.eps)
  )

bar_age_three <- bind_rows(
  apo_stack_age %>% mutate(measure = factor("APO", levels = c("APO","ATE","Ratios"))),
  ate_stack_age %>% mutate(measure = factor("ATE", levels = c("APO","ATE","Ratios"))),
  pt_ratio_df
)

ggplot(bar_age_three, aes(x = age, y = value, fill = component)) +
  geom_col(width = 0.7) +
  facet_grid(rows = vars(measure), cols = vars(gender), scales = "free_y") +
  scale_fill_manual(
    values = c(
      "Truth" = "#1f77b4", "PT bias" = "#fdae6b", "PT / DID APO" = "#99d8c9"
    )
  ) +
  labs(
    title = "Figure 4: Normalizing the PT Bias",
    subtitle = "Bottom row shows PT bias / DID APO",
    x = "Age", y = NULL, fill = NULL
  ) +
  theme(legend.position = "bottom")
```

**Critical observation:** The bottom row (PT / DID APO) is **identical for men and women** at every age. Hence, differencing across gender cancels out this bias parameter. This is an implication of the NTD assumption.

### Step 2: The Complete Ratio

Now let's add the second part of the sum, $\frac{\text{ATE}}{\text{DID APO}}$.

```{r fig5_full_ratio, fig.height=5.5}
ratios_df <- summary_age %>%
  transmute(
    gender, age = factor(age),
    measure = factor("Ratios", levels = c("APO","ATE","Ratios")),
    `PT ratio`  = PT_bias / pmax(APO_true - PT_bias, .Machine$double.eps),
    `ATE ratio` = ATE_true / pmax(APO_true - PT_bias, .Machine$double.eps)
  ) %>%
  pivot_longer(
    cols = c(`PT ratio`, `ATE ratio`), 
    names_to = "component", 
    values_to = "value"
  )

bar_age_three2 <- bind_rows(
  apo_stack_age %>% mutate(measure = factor("APO", levels = c("APO","ATE","Ratios"))),
  ate_stack_age %>% mutate(measure = factor("ATE", levels = c("APO","ATE","Ratios"))),
  ratios_df
)

ggplot(bar_age_three2, aes(x = age, y = value, fill = component)) +
  geom_col(width = 0.7) +
  facet_grid(rows = vars(measure), cols = vars(gender), scales = "free_y") +
  scale_fill_manual(
    values = c(
      "Truth" = "#1f77b4", "PT bias" = "#fdae6b",
      "PT ratio" = "#99d8c9", "ATE ratio" = "#2ca25f"
    )
  ) +
  labs(
    title = "Figure 5: Complete Decomposition of the Normalized Ratio",
    subtitle = "Bottom row: Light teal (drops out when differencing) + Dark green (what remains)",
    x = "Age", y = NULL, fill = NULL
  ) +
  theme(legend.position = "bottom")
```

**The bottom row shows:**

- Light teal (PT / DID APO): Same across gender $\rightarrow$ cancels when we difference between gender
- Dark green (ATE / DID APO): Different across gender $\rightarrow$ now we need to understand, if we difference this between gender, how much is bias? How much is effect?

## The Multiplicative Bias Structure

Here's where it gets interesting. The dark green component (ATE / DID APO) can be rewritten to reveal a multiplicative bias:

$$\frac{\text{ATE}}{\text{DID APO}} = \frac{\text{ATE}}{\text{True APO} - \text{PT bias}} = \theta \cdot \frac{\text{True APO}}{\text{True APO} - \text{PT bias}}$$

where $\theta = \text{True ATE} / \text{True APO}$ is the true ratio we want.

This simplifies to:
$$\frac{\text{ATE}}{\text{DID APO}} = \theta + \theta \cdot \frac{\text{PT bias}}{\text{True APO} - \text{PT bias}}$$

The second term is **multiplicative bias** - the true effect gets scaled by the ratio of PT bias to DID APO. As we discussed above, this quantity $\frac{\text{PT bias}}{\text{True APO} - \text{PT bias}}$ is constant between genders. So both genders treatment effects are scaled by the same factor. 

```{r fig6_multiplicative, fig.height=5.5}
eps <- .Machine$double.eps

theta_df <- summary_age %>%
  transmute(
    gender, age = factor(age),
    APO_true, ATE_true, APO_did, PT_bias,
    theta_true = ATE_true / pmax(APO_true, eps),
    bias_A = PT_bias / pmax(APO_true - PT_bias, eps),
    bias_B = theta_true * (APO_true / pmax(APO_true - PT_bias, eps) - 1)
  )

theta_stack <- bind_rows(
  theta_df %>% transmute(
    gender, age, measure = factor("Ratios", levels = c("APO","ATE","Ratios")),
    component = "Truth", value = theta_true
  ),
  theta_df %>% transmute(
    gender, age, measure = factor("Ratios", levels = c("APO","ATE","Ratios")),
    component = "Bias A", value = bias_A
  ),
  theta_df %>% transmute(
    gender, age, measure = factor("Ratios", levels = c("APO","ATE","Ratios")),
    component = "Bias B", value = bias_B
  )
)

bar_age_three3 <- bind_rows(
  apo_stack_age %>% mutate(measure = factor("APO", levels = c("APO","ATE","Ratios"))),
  ate_stack_age %>% mutate(measure = factor("ATE", levels = c("APO","ATE","Ratios"))),
  theta_stack
)

ggplot(bar_age_three3, aes(x = age, y = value, fill = component)) +
  geom_col(width = 0.7) +
  facet_grid(rows = vars(measure), cols = vars(gender), scales = "free_y") +
  scale_fill_manual(
    values = c(
      "Truth" = "#1f77b4", "PT bias" = "#fdae6b",
      "Bias A" = "#99d8c9", "Bias B" = "#2ca25f"
    ),
    breaks = c("Truth", "PT bias", "Bias A", "Bias B"),
    labels = c("Truth", "PT bias", "PT / DID APO", "Truth × Bias")
  ) +
  labs(
    title = "Figure 6: The Multiplicative Bias Structure",
    subtitle = "Bottom row: Blue (true θ) + Teal (drops out) + Green (multiplicative bias)",
    x = "Age", y = NULL, fill = NULL
  ) +
  theme(legend.position = "bottom")
```

**Bottom row interpretation:**

- **Blue (Truth)**: The true ratio $\theta$ we want to estimate
- **Teal (Bias A)**: PT / DID APO - drops out when differencing across gender
- **Green (Bias B)**: $\theta \times \frac{\text{PT bias}}{\text{True APO} - \text{PT bias}}$ - is not equal between gender, since $\theta$ is allowed to be different between gender $\rightarrow$ hence bias remains when differencing

## Key Takeaways

1. **But normalization creates multiplicative bias**: Even when the identifying assumption holds, the estimated ratio contains bias that scales with the true effect

2. **The bias structure**:
   - Additive part (teal): Cancels when differencing across groups
   - Multiplicative part (green): Remains after differencing
   
3. **Magnitude matters**: The multiplicative bias is $\theta \cdot \frac{\lambda}{1-\lambda}$ where $\lambda = \text{PT bias} / \text{True APO}$. As $\lambda$ increases, the bias increases.

This is the size of PT violations is crucial for interpreting normalized event study estimates.
